Value-Suppressing Uncertainty Maps

### 2017-09-18 paper/main.tex
Just In Case submitted version.

Colors in univariate quantitative color maps should be sufficiently far apart as to be perceptually distinguishable~\cite{ware1988color}, and vary in lightness as well as hue to afford an implicit ordering of value~\cite{borland2007rainbow,rogowitz2001blair}. Different choices of color maps can highlight different features of the data, and should be chosen with care~\cite{rogowitz1996not}. These principles extend to bivariate color maps, with the additional consideration that our perception of color channels lacks \emph{orthogonality}. That is, while we can perceive e.g., the height of a bar in a bar chart independently of its width, a particular property of a color (e.g., its redness, saturation, transparency, etc.) affect how its other properties are perceived \cite{ware2012information,garner1970integrality}. Many color channels such as hue and saturation are therefore perceptually \emph{integral}, which complicates their use in bivariate maps.

@@

In general, the quality of bivariate color maps is a multivariate measure involving consideration of not just the component color channels, but also the interpolation scheme and the color of the surround~\cite{bernard2015survey}. However, even simple bivariate maps can be difficult for a general audience to interpret: due to their additional complexity, Wainer \& Francolini~\cite{wainer1980empirical} reported high levels of error for participants even for ``elementary'' graph reading tasks, and additionally that bivariate legends were more difficult to memorize and internalize than their univariate equivalents. Therefore, bivariate maps in practice are often limited to a small set of output colors (say, a 4x4 matrix as in \figref{fig:example}) \cite{robertson1986generation,trumbo1981theory}.

Given that bivariate maps, for reasons of either practicality or perceptual fidelity, have only a limited budget of outputs, we adapt an insight from Dunn~\cite{dunn1989dynamic} that these limited categories ought to be assigned with regards to their importance to analysis tasks, rather than uniformly. Our Value-Suppressing Uncertainty Palettes apply this insight to data with uncertainty, where, frequently, uncertain information ought to be weighted less heavily in a decision-making process, or at least less prominence in an analysis, than certain information.

@@

When selecting a visual channel to represent value, the designer should consider the perceptual integrality of the value and uncertainty visual channels. For instance, we use the Viridis~\cite{viridis} color map frequently in this paper, as many other standard color ramps (such as sequential ColorBrewer~\cite{harrower2003colorbrewer} ramps), following usual best practices, interpolate in both hue and luminance~\cite{ware1988color}. This interpolation in value interferes with our uncertainty encoding, introducing ambiguity. Ramps such as Viridis, which avoid very light and very dark colors, reduce this ambiguity.

When determining the quantization tree, the design should be careful to avoid having too many output colors. For reasons of memorability and complexity, Wainer\& Francolini~\cite{wainer1980empirical} suggest no more than 16. For a binary quantization tree, this results in a maximum tree depth of 4, as seen in \figref{fig:example}. In general, for tasks that require finer-grade distinctions between uncertainty, the depth of the tree could be increased. For tasks involving finer-grade distinctions between value, the branching factor of the tree could be increased. However, these factors increase the number of colors, and so designers should moderate between legend complexity and data fidelity. Beyond these two parameters, non-uniform binning would allow the designer to target particular distributions or important subregions of the data that would otherwise be difficult to visualize in a limited palette, as in Dunn's~\cite{dunn1989dynamic} dynamic bivariate maps. As an additional form of control, prior tools using VSUP-like techniques, such as LayerCake~\cite{correll2015layercake}, allow the analyst to interactively control the size of bins in the tree through dragging on the color legend.

@@

For the identification task, we gave participants a $5$x$5$ heatmap, and asked them to click on a region of the heatmap that encodes a particular (value, uncertainty) pair. To support comparability across conditions, we chose a set of target pairs such that each target mapped to a unique color across all encoding types. This resulted in (4+4+2+1)=11 valid targets. For each trial, we placed one example of each of these 11 targets randomly in a heatmap; the remaining 13 cells of the heatmap were randomly sampled from this target list.

@@

The results support our first hypothesis. Juxtaposition was a significant effect ($F(1,166)=5.5$,$p=0.02$). A post-hoc t-test confirmed ($t=2.5$, $p=0.01$) that superimposed charts ($M=0.58$, $SD=0.5$) performed significantly better than juxtaposed ($M=0.51$, $SD=0.49$) ones. For our identification task, juxtaposition requires at least two search tasks (one to locate squares with the correct data value, another to locate ones with the correct uncertainty value), which introduces additional error compared to conditions where value and uncertainty are displayed simultaneously. This is reflected also in response times, which, although highly variable, were highest in juxtaposed conditions ($M=19$, $SD=65$), than the other conditions ($M=13$, $SD=38$), although this difference was not significant ($t=1.64$, $p=0.1$).

@@

For the prediction task, we gave participants the rules of a game similar to Battleship. Greis \ea~\cite{greis2016decision} employ these game-like experimental tasks to assess how different visual designs communicate uncertainty information, which can be abstract or complex, to the general audience. In our task, the participant and a (fictional) adversary have to place tokens representing ships on a $5$x$5$ spatial grid, with the expectation that certain squares will be hit by missiles. Players have to place all their tokens before continuing. The objective is to minimize the number of your own ships that are hit. In our task, participants were given a map representing the predictions of missile strikes in each location on the grid. The \emph{value} component of the prediction was the likelihood of a missile strike in a particular region: the ship's \emph{safety} if placed on the square. The \emph{uncertainty} component was the confidence in this prediction. Other studies of uncertainty representation, such as in Cox et al.~\cite{cox2013visualizing}, have used "prediction + prediction uncertainty" stimuli to elicit differences in decision-making between visualizations of uncertainty.

@@

The results partially support our first hypothesis. We performed a repeated measures ANOVA on our results to measure the effect of VSUP versus standard quantizations, and square- versus wedge-shaped legends, on average uncertainty in bets. We found no significant effect of quantization scheme on average guess uncertainty ($F(1,61)=0.05$, $p=0.83$). This result indicates that there does not appear to be a uniform pattern of risk aversion between the two scale types. However, given our quartile-based stimuli design, the central tendency of uncertainty would not necessarily capture caution in guesses. To capture differences in this non-normal distribution, and to test our belief that VSUP users would avoid highly uncertain regions, we performed a one-sided two-sample, Kolmogorov-Smirnov test on the distributions of uncertainty in guesses in the VSUP and traditional bivariate conditions. We found a significant difference ($D = 0.5$, $p=0.03$), and an inspection of the distribution shows that participants using VSUPs were less likely to make guesses with high uncertainty. \figref{fig:taskTwoQuartiles} shows this result.

@@

These two results, taken together, suggest that VSUPs led participants to moderate their guesses: avoiding highly uncertain regions, but willing to ``gamble'' more on regions of middling certainty. In our design, where these gambles were usually poor (due to the predominance of ``bad'' cells), this moderation resulted in lower expected values. However, when value and uncertainty are orthogonal, this moderation can be beneficial, or at least not harmful. Indeed, for guesses with middling and low uncertainty (quartiles 1-3), where this orthogonality was present, there was no significant difference ($t=1.56$, $p=0.11$) in average danger between VSUP ($M=0.30$, $SD=0.24$) and traditional ($M=0.29$, $SD=024$) conditions.

@@

>>>>

Colors in univariate quantitative color maps should be sufficiently far apart as to be perceptually distinguishable~\cite{ware1988color}, and vary in lightness as well as hue to afford an implicit ordering of value~\cite{borland2007rainbow,rogowitz2001blair}. Different choices of color maps can highlight different features of the data, and should be chosen with care~\cite{rogowitz1996not}. These principles extend to bivariate color maps, with the additional consideration that our perception of color channels lacks \emph{orthogonality}. That is, while we can perceive e.g., the height of a bar in a bar chart independently of its width, a particular property of a color (e.g., its redness, saturation, transparency, etc.) affect how its other properties are perceived \cite{garner1970integrality,ware2012information}. Many color channels such as hue and saturation are therefore perceptually \emph{integral}, which complicates their use in bivariate maps.

@@

In general, the quality of bivariate color maps is a multivariate measure involving consideration of not just the component color channels, but also the interpolation scheme and the color of the surround~\cite{bernard2015survey}. However, even simple bivariate maps can be difficult for a general audience to interpret: due to their additional complexity, Wainer \& Francolini~\cite{wainer1980empirical} reported high levels of error for participants even for ``elementary'' graph reading tasks, and additionally that bivariate legends were more difficult to memorize and internalize than their univariate equivalents. Therefore, bivariate maps in practice are often limited to a small set of output colors~\cite{robertson1986generation,trumbo1981theory} (say, a 4x4 matrix as in \figref{fig:example}).

Given that bivariate maps, for reasons of either practicality or perceptual fidelity, have only a limited budget of outputs, we adapt an insight from Dunn~\cite{dunn1989dynamic} that these limited categories ought to be assigned with regards to their importance to analysis tasks, rather than uniformly. Our Value-Suppressing Uncertainty Palettes apply this insight to data with uncertainty. Frequently, uncertain information ought to be weighted less heavily in a decision-making process, or at least have less prominence in an analysis, than certain information.

@@

When selecting a visual channel to represent value, the designer should consider the perceptual integrality of the value and uncertainty visual channels. For instance, we use the Viridis~\cite{viridis} color map frequently in this paper, as many standard color ramps (such as sequential ColorBrewer~\cite{harrower2003colorbrewer} ramps), following usual best practices, interpolate in both hue and luminance~\cite{ware1988color}. This interpolation in luminance interferes with our uncertainty encoding, introducing ambiguity. Ramps such as Viridis, which avoid very light and very dark colors, reduce this ambiguity.

When determining the quantization tree, the design should be careful to avoid having too many output colors. For reasons of memorability and complexity, Wainer \& Francolini~\cite{wainer1980empirical} suggest no more than 16 distinct outputs in bivariate maps. For a binary quantization tree, this results in a maximum tree depth of 4, as seen in \figref{fig:example}. In general, for tasks that require finer-grade distinctions between uncertainty, the depth of the tree could be increased. For tasks involving finer-grade distinctions between value, the branching factor of the tree could be increased. However, these factors increase the number of colors, and so designers should moderate between legend complexity and data fidelity. Beyond these two parameters, non-uniform binning would allow the designer to target particular distributions or important subregions of the data that would otherwise be difficult to visualize in a limited palette, as in Dunn's~\cite{dunn1989dynamic} dynamic bivariate maps. As an additional form of control, prior tools using VSUP-like techniques, such as LayerCake~\cite{correll2015layercake}, allow the analyst to control the bins in the tree through interaction with the color legend.

@@

For the identification task, we gave participants a $5$x$5$ heatmap, and asked them to click on a region of the heatmap that encodes a particular (value, uncertainty) pair. To support comparability across conditions, we chose a set of target pairs such that each target mapped to a unique color across all encoding types. This resulted in (4+4+2+1)=11 valid targets. For each trial, we placed one example of each of these 11 targets randomly in a heatmap; the remaining 13 cells of the heatmap were randomly sampled from this target list. Thus there could be multiple correct locations in each trial, deterring a strategy of looking for ``oddball'' colors.

@@

The results support our first hypothesis. Juxtaposition was a significant effect ($F(1,166)=5.5$,$p=0.02$). A post-hoc t-test confirmed ($t=2.5$, $p=0.01$) that superimposed charts ($M=0.58$, $SD=0.5$) performed significantly better than juxtaposed ($M=0.51$, $SD=0.49$) ones. For our identification task, juxtaposition requires at least two search tasks (one to locate squares with the correct data value, another to locate ones with the correct uncertainty value), which introduces additional error compared to conditions where value and uncertainty are displayed simultaneously. This is reflected also in response times, which, although highly variable, were highest in juxtaposed conditions ($M=19$s, $SD=65$), than the other conditions ($M=13$s, $SD=38$), although this difference was not significant ($t=1.64$, $p=0.1$).

@@

For the prediction task, we gave participants the rules of a game similar to Battleship. Greis \ea~\cite{greis2016decision} employ these game-like experimental tasks to assess how different visual designs communicate uncertainty information, which can be abstract or complex, to the general audience. In our task, the participant and a (fictional) adversary have to place tokens representing ships on a $5$x$5$ spatial grid, with the expectation that certain squares will be hit by missiles. Players have to place all their tokens before continuing. The objective is to minimize the number of your own ships that are hit. In our task, participants were given a map representing the predictions of missile strikes in each location on the grid. The \emph{value} component was the ship's \emph{danger} if placed on the square. The \emph{uncertainty} component was the confidence in this prediction. Other studies of uncertainty representation, such as in Cox et al.~\cite{cox2013visualizing}, have used "prediction + prediction uncertainty" stimuli to elicit differences in decision-making between visualizations of uncertainty.

@@

The results partially support our first hypothesis. We performed a repeated measures ANOVA on our results to measure the effect of VSUP versus standard quantizations, and square- versus wedge-shaped legends, on average uncertainty in bets. We found no significant effect of quantization scheme on average guess uncertainty ($F(1,61)=0.05$, $p=0.83$). This result indicates that there does not appear to be a uniform pattern of risk aversion between the two scale types. However, given our quartile-based stimuli design, the central tendency of uncertainty would not necessarily capture caution in guesses. To capture differences in this non-normal distribution, and to test our belief that VSUP users would avoid highly uncertain regions, we performed a one-sided two-sample, Kolmogorov-Smirnov test on the distributions of uncertainty in guesses in the VSUP and traditional bivariate conditions. We found a significant difference ($D = 0.5$, $p=0.03$), and an inspection of the distribution shows that participants using VSUPs were less likely to make guesses with very high uncertainty. \figref{fig:taskTwoQuartiles} shows this result.

@@

These two results, taken together, suggest that VSUPs led participants to moderate their guesses: avoiding highly uncertain regions, but willing to ``gamble'' more on regions of middling certainty. In our design, where these gambles were usually poor (due to the predominance of ``bad'' cells), this moderation resulted in lower expected values. However, when value and uncertainty are orthogonal, this moderation can be beneficial, or at least not harmful. Indeed, for guesses with middling and low uncertainty (quartiles 1-3), where this orthogonality was present, there was no significant difference ($t=1.56$, $p=0.11$) in average danger between VSUP ($M=0.30$, $SD=0.24$) and traditional ($M=0.29$, $SD=0.24$) conditions.

@@

VSUPs, like other mappings involving quantization, are also highly sensitive to the binning scheme used. Uncertain data with values that fall very close to bin boundaries may end up having large color differences, even though they may be close in value. The comparison to effects tests is therefore imperfect, and a particular item's encoding in a VSUP cannot be used as a proxy for a test of statistical significance. As with other quantized forms of data analysis, including histograms and geospatial binning, analysts and designers should consider the distribution of values before quantizing.
